---
title: "Exploring Data Sets"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring Data Sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, comment = "#>")
suppressPackageStartupMessages(library(sjmisc))
```

Tidying up, transforming and exploring data is an important part of data analysis, and you can manage many common tasks in this process with the [tidyverse](http://tidyverse.org/) or related packages. The **sjmisc**-package fits into this workflow, especially when you work with [labelled data](https://cran.r-project.org/package=sjlabelled), because it offers functions for data transformation and labelled data utility functions. This vignette describes typical steps when beginning with data exploration.

The examples are based on data from the EUROFAMCARE project, a survey on the situation of family carers of older people in Europe. The sample data set `efc` is part of this package. Let us see how the family carer's gender and subjective perception of negative impact of care as well as the cared-for person's dependency are associated with the family carer's quality of life.

```{r message=FALSE}
library(sjmisc)
library(dplyr)
data(efc)
```

## Print frequencies with labels

The first thing that may be of interest is probably the distribution of gender. You can plot frequencies for labelled data with `frq()`. This function requires either a vector or data frame as input and prints the variable label as first line, followed by a frequency-table with values, labels, counts and percentages of the vector.

```{r}
frq(efc$c161sex)
```

## Find variables in a data frame

Next, let’s look at the distribution of gender by the cared-for person's dependency. To compute cross tables, you can use `flat_table()`. It requires the data as first argument, followed by any number of variable names.

But first, we need to know the name of the dependency-variable. This is where `find_var()` comes into play. It searches for variables in a data frame by

1. variable names,
2. variable labels, 
3. value labels 
4. or any combination of these. 

By default, it looks for variable name and labels. The function also supports regex-patterns. By default, `find_var()` returns the column-indices, but you can also print a small "summary"" with the `out`-argument.

```{r}
# find all variables with "dependency" in name or label
find_var(efc, "dependency", out = "table")
```

Variable in column 5, named _e42dep_, is what we are looking for.

## Print crosstables with labels

Now we can look at the distribution of gender by dependency:

```{r}
flat_table(efc, e42dep, c161sex)
```

Since the distribution of male and female carers is skewed, let's see the proportions. To compute crosstables with row or column percentages, use the `margin`-argument:

```{r}
flat_table(efc, e42dep, c161sex, margin = "col")
```

## Recoding variables

Next, we need the negatice impact of care (*neg_c_7*) and want to create three groups: low, middle and high negative impact. We can easily recode and label vectors with `rec()`. This function does not only recode vectors, it also allows direct labelling of categories inside the recode-syntax (this is optional, you can also use the `val.labels`-argument). We now recode *neg_c_7* into a new variable _burden_. The cut-points are a bit arbitrary, for the sake of demonstration.

```{r}
efc$burden <- rec(
  efc$neg_c_7,
  rec = c("min:9=1 [low]; 10:12=2 [moderate]; 13:max=3 [high]; else=NA"),
  var.label = "Subjective burden",
  as.num = FALSE # we want a factor
)
# print frequencies
frq(efc$burden)
```

You can see the variable _burden_ has a variable label ("Subjective burden"), which was set inside `rec()`, as well as three values with labels ("low", "moderate" and "high"). From the lowest value in *neg_c_7* to 9 were recoded into 1, values 10 to 12 into 2 and values 13 to the highest value in *neg_c_7* into 3. All remaining values are set to missing (`else=NA` – for details on the recode-syntax, see `?rec`).

## Grouped data frames

How is burden distributed by gender? We can group the data and print frequencies using `frq()` for this as well, as this function also accepts grouped data frames. Frequencies for grouped data frames first print the group-details (variable name and category), followed by the frequency table. Thanks to labelled data, the output is easy to understand.

```{r}
efc %>% 
  select(burden, c161sex) %>% 
  group_by(c161sex) %>% 
  frq()
```

## Nested data frames

Let's investigate the association between quality of life and burden across the different dependency categories, by fitting linear models for each category of _e42dep_. We can do this using _nested data frames_. `nest()` from the **tidyr**-package can create subsets of a data frame, based on grouping criteria, and create a new _list-variable_, where each element itself is a data frame (so it’s nested, because we have data frames inside a data frame).

In the following example, we group the data by _e42dep_, and "nest" the groups. Now we get a data frame with two columns: First, the grouping variable (_e42dep_) and second, the datasets (subsets) for each country as data frame, stored in the list-variable _data_. The data frames in the subsets (in _data_) all contain the selected variables _burden_, _c161sex_ and *quol_5* (quality of life).

```{r}
# convert variable to labelled factor, because we then 
# have the labels as factor levels in the output
efc$e42dep <- to_label(efc$e42dep, drop.levels = T)
efc %>%
  select(e42dep, burden, c161sex, quol_5) %>%
  group_by(e42dep) %>%
  tidyr::nest()
```

## Get coefficients of nested models

Using `map()` from the **purrr**-package, we can iterate this list and apply any function on each data frame in the list-variable "data". We want to apply the `lm()`-function to the list-variable, to run linear models for all "dependency-datasets". The results of these linear regressions are stored in another list-variable, _models_ (created with `mutate()`). To quickly access and look at the coefficients, we can use `spread_coef()`.

```{r}
efc %>%
  select(e42dep, burden, c161sex, quol_5) %>%
  group_by(e42dep) %>%
  tidyr::nest() %>% 
  na.omit() %>%       # remove nested group for NA
  arrange(e42dep) %>% # arrange by order of levels
  mutate(models = purrr::map(
    data, ~ 
    lm(quol_5 ~ burden + c161sex, data = .))
  ) %>%
  spread_coef(models)
```

We see that higher burden is associated with lower quality of life, for all dependency-groups. The `se` and `p.val`-arguments add standard errors and p-values to the output. `model.term` returns the statistics only for a specific term. If you specify a `model.term`, arguments `se` and `p.val` automatically default to `TRUE`.

```{r}
efc %>%
  select(e42dep, burden, c161sex, quol_5) %>%
  group_by(e42dep) %>%
  tidyr::nest() %>% 
  na.omit() %>%       # remove nested group for NA
  arrange(e42dep) %>% # arrange by order of levels
  mutate(models = purrr::map(
    data, ~ 
    lm(quol_5 ~ burden + c161sex, data = .))
  ) %>%
  spread_coef(models, burden3)
```
