---
title: "Working with Labelled Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Labelled Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Labelled Data

In software like SPSS, it is common to have value and variable labels as variable attributes. Variable values, even if categorical, are mostly numeric. In R, however, you may use labels as values directly:

```{r collapse=TRUE}
factor(c("low", "high", "mid", "high", "low"))
```

Reading SPSS-data (from `haven`, `foreign` or `sjmisc`), keeps the numeric values for variables, but adds the value and variable labels as attributes. See following example from the sample-dataset `efc`, which is part of the **sjmisc**-package:

```{r collapse=TRUE}
library(sjmisc)
data(efc)
str(efc$e42dep)
```

While all plotting and table functions of the [sjPlot-package](http://cran.r-project.org/web/packages/sjPlot/index.html) make use of labelled data, many packages and/or functions do not consider these attributes, e.g. R base graphics:

```{r fig.width=6, fig.height=4, dpi=72, dev="png", warning=FALSE}
library(sjmisc)
data(efc)
barplot(table(efc$e42dep, efc$e16sex), 
        beside = T, 
        legend.text = T)
```

As you can see in the above figure, the plot has neither axis nor legend labels.

# Adding value labels as factor values

`to_label` converts a numeric variable into a factor and sets value labels as factor levels. When using factors with "labelled"" levels, the bar plot will also be labelled.

```{r fig.width=6, fig.height=4, dpi=72, dev="png", warning=FALSE}
barplot(table(to_label(efc$e42dep),
              to_label(efc$e16sex)), 
        beside = T, 
        legend.text = T)
```

# Adding variable names as title

There are four functions that let you easily set or get value and variable labels of either a single vector or a complete data frame (see also other vignettes):

  * `get_label()` to get variable labels
  * `get_labels()` to get value labels
  * `set_label()` to set variable labels (add them as vector attribute)
  * `set_labels()` to set value labels (add them as vector attribute)

You can easily add titles to plots dynamically, i.e. depending on the variable that is plotted.

```{r fig.width=6, fig.height=4, dpi=72, dev="png", warning=FALSE}
barplot(table(to_label(efc$e42dep),
              to_label(efc$e16sex)), 
        beside = T, 
        legend.text = T,
        main = get_label(efc$e42dep))
```

# Another example

Converting `labelled` vectors into `factor`s usually drops label attributes (e.g. using `as_factor`) or replaces values with the associated labels (like `to_label` does). If you want to convert a labelled vector into a numeric factor, but keep the label attributes (including variable labels), use `to_factor`.

Functions like `lm` simply copy these attributes and store these information in the returned object; see following example from the `sjPlot`-package:

```{r collapse=TRUE}
library(sjPlot)
data(efc)
# make education categorical
efc$c172code <- to_factor(efc$c172code)
fit <- lm(barthtot ~ c160age + c12hour + c172code + c161sex, 
          data = efc)
```

```{r eval=FALSE, collapse=TRUE}
sjt.lm(fit, group.pred = TRUE)
```
`r sjt.lm(fit,group.pred = TRUE, no.output=TRUE)$knitr`

Looking at `str(fit$frame)` shows us that both variable and value label attributes are still there. Packages like `sjPlot` make use of this feature and automatically label the table output (like seen above).

# Restore labels from subsetted data

The base `subset` function as well as `dplyr`'s (at least up to 0.4.1) `filter` and `select` functions drop label attributes (or vector attributes in general) when subsetting data. The **sjmisc**-package offers handy functions to deal with this problem: `copy_labels` and `remove_labels`.

`copy_labels` adds back labels to a subsetted data frame based on the original data frame. And `remove_labels` removes all label attributes (this might be necessary when working with `dplyr` up to 0.4.1; `dplyr` sometimes throws an error when working with labelled data – this issue should be fixed as of `dplyr` 0.4.2 and above).


## Losing labels during subset

```{r, collapse=TRUE, message=FALSE}
efc.sub <- subset(efc, 
                  subset = e16sex == 1, 
                  select = c(4:8))
str(efc.sub)
```

## Add back labels

```{r, collapse=TRUE}
efc.sub <- copy_labels(efc.sub, efc)
str(efc.sub)
```

# Conclusion

When working with labelled data, especially when working with data sets imported from other software packages, it comes very handy to make use of the label attributes. The **sjmisc**-package supports this feature and offers useful functions for these tasks.
